---
title: "Random Forest - Predicting Titanic Survival"
output: html_document
---

```{r include=FALSE}
library(randomForest)
library(caret)
library(ROCR)
```

Load in data
```{r}
setwd("C:/Users/elrivas/Documents/Trainings/R Training/Titanic")
total_data <- read.csv("C:/Users/elrivas/Documents/Trainings/R Training/Titanic/train.csv")
```

Setting missing age values to the mean
```{r}
total_data$Age[is.na(total_data$Age)] <- mean(total_data$Age, na.rm=TRUE)
```

Create new variable for family size
```{r}
total_data$fam_size <- (total_data$SibSp+total_data$Parch)+1
```

Factor categorical variables. Tell R that the order of classes goes 3,2,1 not 1,2,3.
```{r}
total_data$Survived <- as.factor(total_data$Survived)
total_data$Embarked <- as.factor(total_data$Embarked)
total_data$Sex <- as.factor(total_data$Sex)
total_data$Pclass <- factor(total_data$Pclass, levels=c(3,2,1), order = TRUE)
```

Function to split data
```{r}
split_function <- function(raw, prop.split){
  # randomize
  raw <- raw[sample(1:nrow(raw)),]
  # new variable, make default "train" for now
  raw$test_or_train <- "train"
  raw$test_or_train[sample(1:nrow(raw), size=((prop.split)*nrow(raw)))] <- "test"
  return(raw)
}
```

Run function and subset data into "test" and "train"
```{r}
new_data <- split_function(total_data, .5)

train <- subset(new_data, test_or_train=="train")
test <- subset(new_data, test_or_train=="test")
```

Get variable names and consider which ones would be helpful for predicting survival. Passenger ID, Name, Ticket, and Cabin won't contribute to the model.
```{r}
names(total_data)
```

Run model. Set seed to ensure reproducability. Error plateaus around 1000 trees. Set trees to 1000.
```{r}
set.seed(500)
rf <- randomForest(Survived ~ Pclass + Sex + Age + fam_size + Fare + Embarked, 
                   data=train, importance=TRUE, ntree=1000)
plot(rf)
```

See which variables are most important in model. Sex and Pclass are most important in predicting death.
```{r}
varImpPlot(rf)
```

Use the rf to predict survival
```{r}
pred <- predict(rf, test)
```

Calculate the accuracy of the model with a confusion matrix
```{r}
accuracy <- confusionMatrix(data=pred, reference = test$Survived)$overall[1]
print(accuracy)
```
